Fusing Filters with Integer Linear Programming

### 2014-05-15 2014betterfusionforfilters/section/01-Introduction.tex
remove those damn figure[H] because they were reordering things strangely
minor change to contribution

To compile subgraphs which are the result of our clustering technique into loops, we use data flow fusion~\cite{lippmeier2013flow}. It improved on existing related array fusion approaches (~\cite{coutts2007streamfusion}, ~\cite{keller2010repa} as it 1) it fuses programs that use branching data flows where a produced array is consumed by several consumers, and 2) guarantees complete fusion into a single loop for all programs that operate on the same size input data and contain no fusion-preventing dependencies between operators.

@@

To demonstrate the effect of clustering and fusion, consider the following code snippet:
\begin{figure}[H]
\begin{code}
 normalize2 :: Array Int -> Array Int
 normalize2 xs
  = let sum1 = fold   (+)  0   xs
        gts  = filter (> 0)    xs
        sum2 = fold   (+)  0   gts
        ys1  = map    (/ sum1) xs
        ys2  = map    (/ sum2) xs
    in (ys1, ys2)
\end{code}
\end{figure}
The function @normalize2@ computes two sums, one of all the elements of @xs@, the other of all the ones greater than zero. Since we need to fully evalute the sums to proceed with the maps, it is clear that we need at least two separate loops. These folds are examples of fusion-preventing dependencies, as fold cannot be fused with subsequent operations. Figure~\ref{f:normalize2-clusterings} shows the data-flow graph of @normalize@. In the leftmost diagram we can see the effect of applying stream fusion to the program: we end up with four loops (denoted by the dotted lines): only the filter operation is combined with the subsequent fold. The best existing ILP approach results in the right-most graph: it combines the @sum1@ fold and @sum2@ filter in one loop, but requires an extra loop for the fold operation which consumes the filter output, since it cannot fuse filter operations. Our approach (diagram in the middle) produces the, for this example, optimal solution: two loops. The first to calculate both sums, the second to calculate the maps.

@@

% NOTE: This set of bullets needs to fit on the first page, without spilling to the second.
Our contributions are as follows:
\begin{itemize}
\item   
We extend prior work by Megiddo~\cite{megiddo1998optimal} and Darte~\cite{darte2002contraction}, with support for size changing operators. Size changing operators can be clustered with the operators that generate their source arrays, and compiled naturally with data-flow fusion (\S\ref{s:ILP}).

>>>>

To compile subgraphs which are the result of our clustering technique into loops, we use data flow fusion~\cite{lippmeier2013flow}. It improved on existing related array fusion approaches (~\cite{coutts2007streamfusion}, ~\cite{keller2010repa} as it 1) fuses programs that use branching data flows where a produced array is consumed by several consumers, and 2) guarantees complete fusion into a single loop for all programs that operate on the same size input data and contain no fusion-preventing dependencies between operators.

@@

To demonstrate the effect of clustering and fusion, consider the following code snippet:
\begin{code}
 normalize2 :: Array Int -> Array Int
 normalize2 xs
  = let sum1 = fold   (+)  0   xs
        gts  = filter (>   0)  xs
        sum2 = fold   (+)  0   gts
        ys1  = map    (/ sum1) xs
        ys2  = map    (/ sum2) xs
    in (ys1, ys2)
\end{code}
The function @normalize2@ computes two sums, one of all the elements of @xs@, the other of all the ones greater than zero. Since we need to fully evalute the sums to proceed with the maps, it is clear that we need at least two separate loops. These folds are examples of fusion-preventing dependencies, as fold cannot be fused with subsequent operations. Figure~\ref{f:normalize2-clusterings} shows the data-flow graph of @normalize2@. In the leftmost diagram we can see the effect of applying stream fusion to the program: we end up with four loops (denoted by the dotted lines): only the filter operation is combined with the subsequent fold. The best existing ILP approach results in the right-most graph: it combines the @sum1@ fold and @sum2@ filter in one loop, but requires an extra loop for the fold operation which consumes the filter output, since it cannot fuse filter operations. Our approach (diagram in the middle) produces, in this case, the optimal solution: two loops. The first to calculate both sums, the second to calculate the maps.

@@

% NOTE: This set of bullets needs to fit on the first page, without spilling to the second.
Our contributions are as follows:
\begin{itemize}
\item   
We extend prior work by Megiddo~\cite{megiddo1998optimal} and Darte~\cite{darte2002contraction}, with support for size changing operators.
Size changing operators can be clustered with operations on both their source array and output array, and compiled naturally with data-flow fusion (\S\ref{s:ILP}).
